Why is it that statistical tests for residuals are not widely used? An explanation using visual inference.

### 2023-02-07 paper_comparison/paper_comparison/paper_comparison.Rmd
update the text in the same section

For non-linearity patterns, we do not observe significant difference in power of visual test between distributions, except for the discrete uniform distribution. It could be due to the fact that humans have difficulties in recognising visual patterns from discreteness, which makes the shape disconnected and incomplete (ref?). In terms of conventional tests, the discrete uniform distribution is instead the one with the greatest power, followed by the lognormal distribution, the uniform distribution and the normal distribution.

For heteroskedasticity patterns, the power of visual test on lineups with uniform distribution has the greatest power. This is as expected since other three distributions could reduce the chance of revealing the underlying visual pattern because of uneven data points. Although the power of visual test under the discrete uniform distribution is the lowest in the case of non-linearity patterns, it has a relatively great power this time. Considering for heteroskedasticity, the visual patterns are usually detected by connecting the maximum and minimum residuals separately at different fitted values. It will not be greatly affected by the discrete uniform distribution compared to the uniform distribution. Visual tests have low power on lineups with the normal distribution and the lognormal distribution. This suggests the use of transformation of fitted values in reading residual plots. In the case of conventional tests, the discrete uniform distribution consistently perform well, followed by the uniform distribution, the normal distribution and the lognormal distribution.

In regards of the type of the simulation model, the power of visual tests under different fitted value distributions are always lower than the corresponding conventional tests, which is consistent with the result reported in the previous sections.

@@

```{r different-x-dist-poly-power, fig.cap = "Comparison of power on lineups with different fitted value distributions, for conventional tests and visual tests. For visual tests, all the power curves are estimated using five evaluations on each lineup. For lineups with uniform fitted value distribution, the five evaluations are randomly sampled from the total eleven evaluations. The sampling process has been repeated for 1000 times and the corresponding power curves are drawn with transparent grey lines. For both types of tests power does change. The discrete fitted value distribution substantially lowers the power of visual tests for non-linearity patterns in residual plots. For heteroskedasticity patterns, the lognormal distribution produce lower power for both types of tests."}

@@

@@

<!-- ### Shape of non-linearity (old) -->

<!-- Shape of the non-linearity pattern is another factor contributed to the non-linearity model. According to Figure \ref{fig:poly-power-j}, conventional tests are significantly less powerful in testing "Triple-U" shape compared to other three shapes, though it is still more sensitive than the visual test. In fact, among the lineups containing non-linearity patterns used in our experiments, around `r vi_survey %>% filter(!attention_check) %>% filter(!null_lineup) %>% filter(type == "polynomial") %>% group_by(unique_lineup_id) %>% summarise(across(everything(), first)) %>% count(a = p_value < 0.05, b = conventional_p_value < 0.05) %>% mutate(prop = n / sum(n) * 100) %>% filter(a, !b) %>% pull(prop) %>% round(digits = 1)`% (two our of 576) are rejected by the visual test, but accepted by the conventional test. -->

<!-- Figure \ref{fig:poly-example} shows one of the two lineups where the visual test rejects the residual plot. The data plot, in position 18, is very obvious, and different from the nulls. The visual test is correct to reject. To understand why this is, one needs to return to the way the RESET test is applied. It requires a parameter indicating degree of fitted values to test for, and the recommendation is to generically use four (@ramsey_tests_1969). However, the "Triple-U" shape constructed from the Hermite polynomials use power up to 18. Both lineups rejected by the visual test but not the RESET are of "Triple-U". If the RESET test had been applied using the power up to six, all the decisions for these lineups would have been to reject. The recommendation of the polynomial power for the RESET should be revised, perhaps. This illustrates the sensitivity of conventional testing to the parameters, and it also points to a limitation that one needs to know the data structure in order to set the parameters for the test. -->

<!-- We expect the "U" shape will be the easiest one to be detected by subjects followed by the "S", "M" and "Triple-U" shape. From Figure \ref{fig:poly-power-j}, it can be observed that the power curves are mostly aligned with our expectation, except for the "M" shape. Interestingly, the "M" shape is as easy to be detected as the "U" shape. The readability of the shape do not strictly follow the degree of the polynomials. -->

<!-- ```{r poly-power-j, fig.height = 4, fig.cap = "Power of conventional tests and visual tests on lineups containing four different non-linearity shapes. Power curves with higher order of non-linearity are drawn with deeper colours."} -->

<!-- poly_conv_sim %>% -->
<!--   filter(x_dist == "uniform") %>% -->
<!--   mutate(log_effect_size = log(effect_size)) %>% -->
<!--   mutate(reject = as.numeric(RESET4_p_value < 0.05)) %>% -->
<!--   mutate(shape = c("U", "S", "M", "Triple-U")[shape]) %>% -->
<!--   mutate(conv_or_not = "Conventional") %>% -->
<!--   bind_rows(vi_survey %>% -->
<!--   filter(!attention_check) %>% -->
<!--   filter(!null_lineup) %>% -->
<!--   filter(x_dist == "uniform", type == "polynomial") %>% -->
<!--   mutate(log_effect_size = log(effect_size)) %>% -->
<!--   mutate(reject = as.numeric(p_value < 0.05)) %>% -->
<!--   mutate(shape = c("U", "S", "M", "Triple-U")[shape]) %>% -->
<!--   mutate(conv_or_not = "Visual")) %>% -->
<!--   mutate(shape = factor(shape, levels = c("U", "S", "M", "Triple-U"))) %>% -->
<!--   select(shape, log_effect_size, reject, conv_or_not) %>% -->
<!--   ggplot() + -->
<!--   geom_smooth(aes(log_effect_size, reject, col = shape),  -->
<!--               se = FALSE,  -->
<!--               method = "glm", -->
<!--               method.args = list(family = binomial())) + -->
<!--   xlab(expression(log[e] (Effect_size))) + -->
<!--   ylab("Power") + -->
<!--   scale_color_manual(values = rcartocolor::carto_pal(4, "TealGrn")) + -->
<!--   theme_light() + -->
<!--   theme(legend.position = "bottom") + -->
<!--   facet_wrap(~conv_or_not) + -->
<!--   labs(col = "Non-linearity shape") -->
<!-- ``` -->

<!-- ```{r poly-example, fig.height = 7, fig.width = 7, fig.cap = "Example of a non-linearity that is rejected by the visual test but not by the RESET test. The data plot at panel 18 exhibits strong non-linear pattern. It has effect size $= 3.96$. Interestingly, all the lineups in this group are higher order polynomials (Triple-U), and it is apparent that the default RESET test under-performs. The RESET test using the higher order polynomial parameter could be used, but this would change it's performance for lower order polynomials. The only way for it to be sensitive to these type of departure is to know the order prior to testing!"} -->
<!-- # plot poly-275 -->
<!-- VI_MODEL$plot_lineup(vi_lineup$poly_275$data,  -->
<!--                      remove_grid_line = TRUE,  -->
<!--                      theme = theme_light(), -->
<!--                      remove_axis = TRUE) -->
<!-- ``` -->

<!-- ### Shape of heteroskedasticity (old) -->

<!-- We also investigate the impact of the shape of heteroskedasticity on the power of different tests. The corresponding result is presented in Figure \ref{fig:heter-power-a}. In theory, the "Left-triangle" and the "Right-triangle" should have little difference from the point of view of a Breusch-Pagan test. According to Figure \ref{fig:heter-power-a}, this is indeed the case. And these two shapes should have the same visual test power, unless the orientation of the visual pattern is a non-negligible factor from human perspective. However, by checking the bootstrap confidence intervals, it can be observed from Figure \ref{fig:heter-power-a} that when $3 < log(E) < 4$, the "Left-triangle" power curve is significantly greater than the "Right-triangle" power curve. This phenomenon is particularly interesting. It implies the current residual plot design is not robust to the orientation of the visual pattern, which is worth to be explored in future research.  -->

<!-- ```{r heter-power-a, fig.height = 4, fig.cap = "Power of conventional tests and visual tests on lineups produced by the heteroskedasticity model with three different shapes controlled by the parameter $a$."} -->

<!-- heter_conv_sim %>% -->
<!--   filter(x_dist == "uniform") %>% -->
<!--   mutate(log_effect_size = log(effect_size)) %>% -->
<!--   mutate(reject = as.numeric(BP_p_value < 0.05)) %>% -->
<!--   mutate(a = c("Left-triangle", "Butterfly", "Right-triangle")[a + 2]) %>% -->
<!--   mutate(conv_or_not = "Conventional") %>% -->
<!--   bind_rows(vi_survey %>% -->
<!--   filter(!attention_check) %>% -->
<!--   filter(!null_lineup) %>% -->
<!--   filter(x_dist == "uniform", type == "heteroskedasticity") %>% -->
<!--   mutate(log_effect_size = log(effect_size)) %>% -->
<!--   mutate(reject = as.numeric(p_value < 0.05)) %>% -->
<!--   mutate(a = c("Left-triangle", "Butterfly", "Right-triangle")[a + 2]) %>% -->
<!--   mutate(conv_or_not = "Visual")) %>% -->
<!--   mutate(a = factor(a, levels = c("Left-triangle", "Butterfly", "Right-triangle"))) %>% -->
<!--   select(a, log_effect_size, reject, conv_or_not) %>% -->
<!--   ggplot() + -->
<!--   geom_smooth(aes(log_effect_size, reject, col = a),  -->
<!--               se = FALSE,  -->
<!--               method = "glm", -->
<!--               method.args = list(family = binomial())) + -->

<!--   # Draw bootstrap for Left-triangle -->
<!--   geom_ribbon(data = ~get_boot_conf(filter(.x,  -->
<!--                                            conv_or_not == "Visual",  -->
<!--                                            a == "Left-triangle"), -->
<!--                                     log_effect_size, -->
<!--                                     reject, -->
<!--                                     times = 1000) %>% -->
<!--                 mutate(conv_or_not = "Visual"), -->
<!--               aes(x = log_effect_size,  -->
<!--                   ymin = lower,  -->
<!--                   ymax = upper), -->
<!--               fill = rcartocolor::carto_pal(3, "Earth")[1], -->
<!--               alpha = 0.2) + -->

<!--   # Draw bootstrap for Right-triangle -->
<!--   geom_ribbon(data = ~get_boot_conf(filter(.x,  -->
<!--                                            conv_or_not == "Visual",  -->
<!--                                            a == "Right-triangle"), -->
<!--                                     log_effect_size, -->
<!--                                     reject, -->
<!--                                     times = 1000) %>% -->
<!--                 mutate(conv_or_not = "Visual"), -->
<!--               aes(x = log_effect_size,  -->
<!--                   ymin = lower,  -->
<!--                   ymax = upper), -->
<!--               fill = rcartocolor::carto_pal(3, "Earth")[3], -->
<!--               alpha = 0.2) + -->

<!--   # Theme -->
<!--   xlab(expression(log[e] (Effect_size))) + -->
<!--   ylab("Power") + -->
<!--   scale_color_manual(values = rcartocolor::carto_pal(3, "Earth")) + -->
<!--   theme_light() + -->
<!--   theme(legend.position = "bottom") + -->
<!--   facet_wrap(~conv_or_not) + -->
<!--   labs(col = "Heteroskedasticity shape") -->
<!-- ``` -->

>>>>

For non-linearity patterns, we do not observe significant difference in power of visual test between distributions, except for the discrete uniform distribution. It could be due to the fact that humans have difficulties in recognising visual patterns from discreteness, which makes the shape disconnected and incomplete (ref?). In terms of conventional tests, the discrete uniform distribution is instead the one with the greatest power, followed by lognormal, uniform and the normal distribution.

For heteroskedasticity patterns, the power of visual test on lineups with uniform distribution has the greatest power. This is as expected since other three distributions could reduce the chance of revealing the underlying visual pattern because of uneven data points. Although the power of visual test under the discrete uniform distribution is the lowest in the case of non-linearity patterns, it has a relatively great power this time. Considering for heteroskedasticity, the visual patterns are usually detected by connecting the maximum and minimum residuals separately at different fitted values. It will not be greatly affected by the discrete uniform distribution compared to the uniform distribution. Visual tests have low power on lineups with the normal distribution and the lognormal distribution. This suggests the use of transformation of fitted values in reading residual plots. In the case of conventional tests, the discrete uniform distribution consistently perform well, followed by uniform, normal and the lognormal distribution.

In regards of the type of the simulation model, the power of visual tests under different fitted value distributions are substantially lower than the corresponding conventional tests, which is consistent with the result reported in the previous sections.

@@

```{r different-x-dist-poly-power, fig.cap = "Comparison of power on lineups with different fitted value distributions, for conventional tests and visual tests. For visual tests, all the power curves are estimated using five evaluations on each lineup. For lineups with uniform fitted value distribution, the five evaluations are randomly sampled from the total eleven evaluations. The sampling process has been repeated for 1000 times and the corresponding power curves are drawn with transparent grey lines. The vetical dotted line is drawn to indicate where the visual pattern start to become recognisable. For both types of tests power does change. The discrete fitted value distribution substantially lowers the power of visual tests for non-linearity patterns in residual plots. For heteroskedasticity patterns, the lognormal distribution produce lower power for both types of tests."}

@@

  geom_vline(xintercept = 2, linetype = 2) +

@@

### 2023-02-07 paper_comparison/paper_comparison/paper_comparison.tex
update the text in the same section

\caption{Comparison of power on lineups with different fitted value distributions, for conventional tests and visual tests. For visual tests, all the power curves are estimated using five evaluations on each lineup. For lineups with uniform fitted value distribution, the five evaluations are randomly sampled from the total eleven evaluations. The sampling process has been repeated for 1000 times and the corresponding power curves are drawn with transparent grey lines. For both types of tests power does change. The discrete fitted value distribution substantially lowers the power of visual tests for non-linearity patterns in residual plots. For heteroskedasticity patterns, normal and lognormal distributions produce lower power for both types of tests.}\label{fig:different-x-dist-poly-power}
\end{figure}

>>>>

\caption{Comparison of power on lineups with different fitted value distributions, for conventional tests and visual tests. For visual tests, all the power curves are estimated using five evaluations on each lineup. For lineups with uniform fitted value distribution, the five evaluations are randomly sampled from the total eleven evaluations. The sampling process has been repeated for 1000 times and the corresponding power curves are drawn with transparent grey lines. The vetical dotted line is drawn to indicate where the visual pattern start to become recognisable. For both types of tests power does change. The discrete fitted value distribution substantially lowers the power of visual tests for non-linearity patterns in residual plots. For heteroskedasticity patterns, the lognormal distribution produce lower power for both types of tests.}\label{fig:different-x-dist-poly-power}
\end{figure}
